One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-19 paper/model.tex
minor updates

\section{Governing equations and numerical models}\label{model} 
We consider a self-gravitating, inviscid fluid disc of mass $M_d$ 
orbting a central star of mass $M_*$. We will mainly examine 2D
(or razor-thin) discs in favour of numerical 
resolution, but have also carried out some 3D disc
simulations. Hence, for generality we describe the system in
3D, using both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the 
fluid velocity (the angular velocity being $\Omega \equiv v_\phi/R$), 
$p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We impose a locally isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed $c_s$ is given by 
\begin{align}\label{sound-speed}
  % c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
  c_s(R) = c_{s0}\left(\frac{R}{R_0}\right)^{-q/2}
\end{align}
where $c_{s0} = h R_0\Omega_k(R_0)$ and 
$h$ is the disc aspect-ratio at the reference radius $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient (since, for an ideal gas the
temperature is proportional to $c_s^2$). The vertical disc scale-height is
defined by $H=c_s/\Omega_k$. Thus, a strictly isothermal disc with
$q=0$ has $H\propto R^{3/2}$, and $q=1$ corresponds to a 
a disc with constant $H/R$.

>>>>

\section{Governing equations and numerical models}\label{model} 
We consider a self-gravitating, inviscid fluid disc of mass $M_d$ 
orbiting a central star of mass $M_*$. We will mainly examine 2D
(or razor-thin) discs in favour of numerical 
resolution, but have also carried out some 3D disc
simulations. Hence, for generality we describe the system in
3D, using both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the 
fluid velocity (the angular velocity being $\Omega \equiv v_\phi/R$), 
$p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We impose a locally isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed $c_s$ is given by 
\begin{align}\label{sound-speed}
  % c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
  c_s(R) = c_{s0}\left(\frac{R}{R_0}\right)^{-q/2}
\end{align}
where $c_{s0} = h R_0\Omega_k(R_0)$ and 
$h$ is the disc aspect-ratio at the reference radius $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient (since, for an ideal gas the
temperature is proportional to $c_s^2$). The vertical disc scale-height is
defined by $H=c_s/\Omega_k$. Thus, a strictly isothermal disc with
$q=0$ has $H\propto R^{3/2}$, and $q=1$ corresponds to a 
a disc with constant $H/R$.

### 2014-11-19 paper/paper.tex
minor updates

\maketitle
\begin{abstract} 
  We present numerical simulations of astrophysical discs with an 
  imposed temperature profile and in which a sub-portion of it is
  self-gravitating. This might represent, for example, mass  
  built-up in a circumstellar or protoplanetary
  disc at the outer edge of a `dead zone', in which there
  is inefficient magneto-hydrodynamic turbulence to enable accretion. 
  We conduct numerical gravito-hydrodynamic experiments in 2D using
  the FARGO code and in 3D with the ZEUS-MP and PLUTO codes. In such
  discs we observe the development of a 
  spiral mode with azimuthal wavenumber $m=1$ and co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly stationary 
  one-armed spiral pattern in the self-gravitating region of the
  disc. The development of this one-armed 
  spiral does not require the movement of  
  the central star, as found in previous studies. Instead, we find it 
  is associated with the imposed 
  temperature gradient and the radial disc structure. 
  We apply results from linear theory and attribute the
  growth of this spiral to the non-conservation of angular momentum
  for non-axisymmetric linear perturbations in discs with a forced temperature
  gradient. This may lead to an unstable exchange of angular momentum between the spiral 
  and the background disc. However, because the interaction does
  not explicitly depend on self-gravity, we suggest it may also
  affect low-frequency one-armed spirals in non-self-gravitating
  discs. 
\end{abstract}

@@

% but also natural in GI.
However, spiral arms are also associated with global gravitational 
instability (GI) of differentially rotating discs
\citep{goldreich65,laughlin96b,laughlin98,nelson98,lodato05,forgan11}. Large-scale
spiral arms can provide significant angular momentum transport
neccessary for accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
spiral structures due to GI are potentially observable with the Atacama Large 
Millimeter/sub-millimeter Array \citep{cossins10,dipierror14}.

% when/where does it happen? formation stage/dead zones (special for ppd)
GI can be expected in the earliest stages of 
circumstellar disc formation
\citep{kratter10b,inutsuka10,tsukamoto13}, and may be possible in the
outer parts of the disc \citep{rafikov05,matzner05,kimura12}.  
% outer parts  
In the case of protoplanetary discs, which have complex structure
\citep{armitage10}, GI can also occur in `dead
zones' --- sub-portions of the disc where the magneto-rotational
instability is ineffective for angular momentum transport
\citep{gammie96,turner08,landry13} and therefore has reduced mass
accretion. Dead zones may then become
self-gravitating with sufficient mass built-up
\citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}.

>>>>

\maketitle
\begin{abstract} 
  We present numerical simulations of astrophysical discs with an 
  imposed temperature profile and in which a sub-portion of it is
  self-gravitating. This might represent, for example, mass  
  built-up in a circumstellar or protoplanetary
  disc at the outer edge of a `dead zone', in which there
  is inefficient magneto-hydrodynamic turbulence to enable accretion. 
  We conduct numerical gravito-hydrodynamic experiments in 2D using
  the FARGO code and in 3D with the ZEUS-MP and PLUTO codes. In such
  discs we observe the development of a 
  spiral mode with azimuthal wavenumber $m=1$ and co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly stationary 
  one-armed spiral pattern in the self-gravitating region of the
  disc. The development of this one-armed 
  spiral does not require the movement of  
  the central star, as found in previous studies. Instead, we find it
  to be associated with the imposed 
  temperature gradient and the radial disc structure. 
  We apply results from linear theory and attribute the
  growth of this spiral mode to the non-conservation of angular momentum
  for non-axisymmetric linear perturbations in discs with a forced temperature
  gradient. This may lead to an unstable exchange of angular momentum
  between the spiral mode 
  and the background disc. However, because the interaction does
  not explicitly depend on self-gravity, we suggest it may also
  affect low-frequency one-armed spirals in non-self-gravitating
  discs. 
\end{abstract}

@@

% but also natural in GI.
However, spiral arms are also associated with global gravitational 
instability (GI) of differentially rotating discs
\citep{goldreich65,laughlin96b,laughlin98,nelson98,lodato05,forgan11}. Large-scale
spiral arms can provide significant angular momentum transport
neccessary for mass accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
spiral structures due to GI are potentially observable with the Atacama Large 
Millimeter/sub-millimeter Array \citep{cossins10,dipierror14}.

% when/where does it happen? formation stage/dead zones (special for ppd)
GI can be expected in the earliest stages of 
circumstellar disc formation
\citep{kratter10b,inutsuka10,tsukamoto13}, and may be possible in the
outer parts of the disc \citep{rafikov05,matzner05,kimura12}.  
% outer parts  
In the case of protoplanetary discs, which have complex structure
\citep{armitage10}, GI can also occur in `dead
zones' --- sub-portions of the disc where the magneto-rotational
instability is ineffective for angular momentum transport
\citep{gammie96,turner08,landry13} and therefore has a reduced mass
accretion rate. Dead zones may then become
self-gravitating with sufficient mass built-up
\citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}.
